---
title: Predator communities with invasive pike are associated with changes in parental care and brain gene expression in wild Alaskan stickleback
author: "E. Arredondo, B. Byrd, L.R. Stein"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---
# Behavioral data

## Overview

This section of code analyzes parental care behaviors performed by stickleback males both before and after exposure to both native and invasive predator models, across lakes with and without invasive pike.

## Loading necessary packages and creating themes and labels

```{r}

##Clear environment
rm(list = ls()); gc()

# packages
library(librarian)
shelf(tidyverse, ggplot2, tibble, dplyr, tidyr, knitr, kableExtra, broom, broom.mixed, lme4, lmerTest, emmeans, car, performance, MuMIn, effectsize, sjPlot, ggeffects)


# theme / labels / colors
theme_paper <- function() {
  theme_classic(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
      axis.title = element_text(face = "bold", size = 14),
      axis.text = element_text(size = 12, color = "black"),
      legend.title = element_text(face = "bold", size = 13),
      legend.text = element_text(size = 12),
      strip.background = element_rect(fill = "grey95", color = "grey70"),
      strip.text = element_text(face = "bold", size = 12)
    )
}

# Char = orange, Pike = green, Trout = darker salmon
pal_treatment <- c(
  "Char"  = "#E69F00",
  "Pike"  = "#009E73",
  "Trout" = "#D55E00"
)

pal_lake <- c(
  "Native predator only lakes" = "#0072B2",
  "Native + Invasive predator lakes" = "#CC79A7"
)

lake_label_map <- c(
  "No Pike" = "Native predator only lakes",
  "Pike"    = "Native + Invasive predator lakes"
)

treatment_shape_map <- c(
  "Char"  = 16,
  "Pike"  = 17,
  "Trout" = 15
)


# theme / labels / colors
theme_paper <- function() {
  theme_classic(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
      axis.title = element_text(face = "bold", size = 14),
      axis.text = element_text(size = 12, color = "black"),
      legend.title = element_text(face = "bold", size = 13),
      legend.text = element_text(size = 12),
      strip.background = element_rect(fill = "grey95", color = "grey70"),
      strip.text = element_text(face = "bold", size = 12)
    )
}

# Char = orange, Pike = green, Trout = darker salmon
pal_treatment <- c(
  "Char"  = "#E69F00",
  "Pike"  = "#009E73",
  "Trout" = "#D55E00"
)

pal_lake <- c(
  "Native predator only lakes" = "#0072B2",
  "Native + Invasive predator lakes" = "#CC79A7"
)

lake_label_map <- c(
  "No Pike" = "Native predator only lakes",
  "Pike"    = "Native + Invasive predator lakes"
)

treatment_shape_map <- c(
  "Char"  = 16,
  "Pike"  = 17,
  "Trout" = 15
)

```

## Entering and cleaning data

```{r eval=FALSE, include=FALSE}

#loading data using specific path

behavior <- read.csv("C:/Users/Stei0031/OneDrive - University of Oklahoma/Desktop/Lab Members/Graduate Students/Eric Arredondo/AK Summer 2024/ICB Paper Revision/AK2024_Data.csv")

```

```{r}
  
# load in behavior data
behavior <- read.csv("AK2024_Data.csv")

```

```{r}

#Cleaning behavior data

behavior <- behavior %>%
  dplyr::mutate(
    ID = trimws(as.character(ID)),
    Lake = trimws(as.character(Lake)),
    Stage = trimws(as.character(Stage)),
    Treatment = trimws(as.character(Treatment)),
    Lake_type = trimws(as.character(Lake_type)),
    Ecotype = trimws(as.character(Ecotype)),
    Embryos_stage = trimws(as.character(Embryos_stage)),
    Fanning_time = suppressWarnings(as.numeric(Fanning_time)),
    Bodycondition = suppressWarnings(as.numeric(Bodycondition)),
    Nest_depth = suppressWarnings(as.numeric(Nest_depth)),
    Percent_cover = suppressWarnings(as.numeric(Percent_cover)),
    Latency_return = suppressWarnings(as.numeric(Latency_return))
  ) %>%
  dplyr::mutate(
    Lake_type = dplyr::case_when(
      Lake_type %in% c("No Pike", "No pike", "no pike") ~ "No Pike",
      Lake_type %in% c("Pike", "pike") ~ "Pike",
      TRUE ~ Lake_type
    ),
    Treatment = dplyr::case_when(
      Treatment %in% c("Char", "char") ~ "Char",
      Treatment %in% c("Pike", "pike") ~ "Pike",
      Treatment %in% c("Trout", "trout") ~ "Trout",
      TRUE ~ Treatment
    ),
    Ecotype = dplyr::na_if(Ecotype, ""),
    Embryos_stage = dplyr::na_if(Embryos_stage, "")
  )

print(
  behavior %>%
    dplyr::count(Lake_type, Ecotype, Treatment, Stage, name = "count") %>%
    dplyr::arrange(Stage, Lake_type, Ecotype, Treatment)
)

#We are interested in parenting behavior, so remove males that were fanning but did not have eggs in the nest

behavior %>%
  dplyr::filter(!is.na(Embryos_stage), Embryos_stage != "None") %>%
  droplevels() -> behavior

```

## Behaviora data inspection

Inspecting the distribution of time fanning, latency to return to the nest, difference in time fanning (after - before model predator exposure), including sample sizes.

```{r}

# raw fanning histogram
p_raw_fanning_hist <- ggplot(
  behavior %>% dplyr::filter(!is.na(Fanning_time)),
  aes(x = Fanning_time)
) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of raw fanning time",
    x = "Fanning time (s)",
    y = "Count"
  ) +
  theme_paper()

print(p_raw_fanning_hist)

# raw latency to return histogram
p_raw_return_hist <- ggplot(
  behavior %>% dplyr::filter(!is.na(Latency_return)),
  aes(x = Latency_return)
) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of raw latency to return",
    x = "Latency to return (s)",
    y = "Count"
  ) +
  theme_paper()

print(p_raw_return_hist)


# one row per ID for ID-level covariates
id_info <- behavior %>%
  dplyr::group_by(ID) %>%
  dplyr::summarise(
    Lake_type = dplyr::first(Lake_type[!is.na(Lake_type) & Lake_type != ""]),
    Ecotype = dplyr::first(Ecotype[!is.na(Ecotype) & Ecotype != ""]),
    Treatment = dplyr::first(Treatment[!is.na(Treatment) & Treatment != ""]),
    Lake = dplyr::first(Lake[!is.na(Lake) & Lake != ""]),
    Embryos_stage = dplyr::first(Embryos_stage[!is.na(Embryos_stage) & Embryos_stage != ""]),
    Bodycondition = dplyr::first(Bodycondition[!is.na(Bodycondition)]),
    Nest_depth = dplyr::first(Nest_depth[!is.na(Nest_depth)]),
    Percent_cover = dplyr::first(Percent_cover[!is.na(Percent_cover)]),
    Latency_to_return = dplyr::first(Latency_return[Stage == "Predator" & !is.na(Latency_return)]),
    .groups = "drop"
  )

# Create difference in fanning in dataframe
Fan_Dif <- behavior %>%
  tidyr::pivot_wider(
    id_cols = ID,
    names_from = Stage,
    values_from = Fanning_time
  ) %>%
  dplyr::left_join(id_info, by = "ID") %>%
  dplyr::filter(!is.na(Baseline), !is.na(Predator)) %>%
  dplyr::mutate(
    Fan_dif = Predator - Baseline,
    Embryos_stage_simple = dplyr::case_when(
      Embryos_stage %in% c("Fry", "Mixed") ~ "Other",
      is.na(Embryos_stage) ~ NA_character_,
      TRUE ~ Embryos_stage
    ),
    Lake_type_label = dplyr::recode(Lake_type, !!!lake_label_map)
  ) %>%
  dplyr::mutate(
    Lake_type = factor(Lake_type, levels = c("No Pike", "Pike")),
    Lake_type_label = factor(
      Lake_type_label,
      levels = c("Native predator only lakes", "Native + Invasive predator lakes")
    ),
    Ecotype = factor(Ecotype),
    Treatment = factor(Treatment, levels = c("Char", "Pike", "Trout")),
    Embryos_stage_simple = factor(Embryos_stage_simple),
    Lake = factor(Lake)
  )

print(
  Fan_Dif %>%
    dplyr::count(Lake_type, Ecotype, Treatment, name = "count") %>%
    dplyr::arrange(Lake_type, Ecotype, Treatment)
)

# histogram of Fan_dif
p_fandif_hist <- ggplot(
  Fan_Dif %>% dplyr::filter(!is.na(Fan_dif)),
  aes(x = Fan_dif)
) +
  geom_histogram(bins = 20, fill = "grey70", color = "black") +
  labs(
    title = "Histogram of change in fanning time",
    x = "Predator-induced change in fanning time (s)",
    y = "Count"
  ) +
  theme_paper()

print(p_fandif_hist)

# egg-only behavior dataset
aicc_data <- Fan_Dif %>%
  dplyr::filter(
    !is.na(Fan_dif),
    !is.na(Treatment),
    !is.na(Lake_type),
    !is.na(Ecotype),
    !is.na(Lake),
    !is.na(Bodycondition),
    !is.na(Nest_depth),
    !is.na(Percent_cover),
    !is.na(Embryos_stage_simple),
    Embryos_stage_simple != "None"
  ) %>%
  droplevels()

# sample sizes
nrow(aicc_data)
table(aicc_data$Embryos_stage_simple)
table(aicc_data$Lake_type, aicc_data$Treatment)
table(aicc_data$Ecotype)
table(aicc_data$Ecotype, aicc_data$Lake_type)
table(aicc_data$Ecotype, aicc_data$Treatment)

```

## Checking effect of ecotype

```{r}


# Quick ecotype models


# simple model: does ecotype predict Fan_dif?
eco_lm1 <- lm(Fan_dif ~ Ecotype, data = aicc_data)
summary(eco_lm1)
anova(eco_lm1)

# ecotype + treatment
eco_lm2 <- lm(Fan_dif ~ Ecotype + Treatment, data = aicc_data)
summary(eco_lm2)
anova(eco_lm2)

# ecotype + lake type + treatment
eco_lm3 <- lm(Fan_dif ~ Ecotype + Lake_type + Treatment, data = aicc_data)
summary(eco_lm3)
anova(eco_lm3)

# mixed model version with lake as random intercept
eco_lmer1 <- lmerTest::lmer(Fan_dif ~ Ecotype + (1 | Lake), data = aicc_data)
summary(eco_lmer1)

eco_lmer2 <- lmerTest::lmer(Fan_dif ~ Ecotype + Treatment + (1 | Lake), data = aicc_data)
summary(eco_lmer2)

eco_lmer3 <- lmerTest::lmer(Fan_dif ~ Ecotype + Lake_type + Treatment + (1 | Lake), data = aicc_data)
summary(eco_lmer3)

# compare quick models
AIC(eco_lm1, eco_lm2, eco_lm3, eco_lmer1, eco_lmer2, eco_lmer3)


# Quick ecotype plots


# boxplot: Fan_dif by ecotype
p_ecotype_box <- ggplot(
  aicc_data,
  aes(x = Ecotype, y = Fan_dif)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.shape = NA, fill = "grey80", color = "black") +
  geom_jitter(width = 0.12, size = 2.5, alpha = 0.8) +
  labs(
    title = "Change in fanning time by ecotype",
    x = "Ecotype",
    y = "change in fanning time (s)"
  ) +
  theme_paper()

print(p_ecotype_box)

# boxplot: Fan_dif by ecotype, colored by treatment
p_ecotype_treatment_box <- ggplot(
  aicc_data,
  aes(x = Ecotype, y = Fan_dif, color = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(
    aes(shape = Treatment),
    position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.75),
    size = 2.5,
    alpha = 0.8
  ) +
  scale_color_manual(values = pal_treatment) +
  scale_shape_manual(values = treatment_shape_map) +
  labs(
    title = "Change in fanning time by ecotype and treatment",
    x = "Ecotype",
    y = "Predator-induced change in fanning time (s)",
    color = "Treatment",
    shape = "Treatment"
  ) +
  theme_paper()

print(p_ecotype_treatment_box)

# boxplot: Fan_dif by ecotype, faceted by lake type
p_ecotype_laketype_box <- ggplot(
  aicc_data,
  aes(x = Ecotype, y = Fan_dif, color = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(
    aes(shape = Treatment),
    position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.75),
    size = 2.5,
    alpha = 0.8
  ) +
  scale_color_manual(values = pal_treatment) +
  scale_shape_manual(values = treatment_shape_map) +
  facet_wrap(~ Lake_type_label) +
  labs(
    title = "Change in fanning time by ecotype, treatment, and lake type",
    x = "Ecotype",
    y = "Predator-induced change in fanning time (s)",
    color = "Treatment",
    shape = "Treatment"
  ) +
  theme_paper()

print(p_ecotype_laketype_box)

# mixed model version with lake as random intercept
base_lmer1 <- lmerTest::lmer(Baseline ~ Ecotype + (1 | Lake), data = aicc_data)
summary(base_lmer1)

base_lmer2 <- lmerTest::lmer(Baseline ~ Ecotype + Treatment + (1 | Lake), data = aicc_data)
summary(base_lmer2)

base_lmer3 <- lmerTest::lmer(Baseline ~ Ecotype + Lake_type + Treatment + (1 | Lake), data = aicc_data)
summary(base_lmer3)

# compare quick models
AIC(eco_lm1, eco_lm2, eco_lm3, eco_lmer1, eco_lmer2, eco_lmer3)


# Quick ecotype plots


# boxplot: Fan_dif by ecotype
p_ecotype_box <- ggplot(
  aicc_data,
  aes(x = Ecotype, y = Fan_dif)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.shape = NA, fill = "grey80", color = "black") +
  geom_jitter(width = 0.12, size = 2.5, alpha = 0.8) +
  labs(
    title = "Change in fanning time by ecotype",
    x = "Ecotype",
    y = "change in fanning time (s)"
  ) +
  theme_paper()

print(p_ecotype_box)

# boxplot: Fan_dif by ecotype, colored by treatment
p_ecotype_treatment_box <- ggplot(
  aicc_data,
  aes(x = Ecotype, y = Fan_dif, color = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(
    aes(shape = Treatment),
    position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.75),
    size = 2.5,
    alpha = 0.8
  ) +
  scale_color_manual(values = pal_treatment) +
  scale_shape_manual(values = treatment_shape_map) +
  labs(
    title = "Change in fanning time by ecotype and treatment",
    x = "Ecotype",
    y = "Predator-induced change in fanning time (s)",
    color = "Treatment",
    shape = "Treatment"
  ) +
  theme_paper()

print(p_ecotype_treatment_box)

# boxplot: Fan_dif by ecotype, faceted by lake type
baseline_ecotype_laketype_box <- ggplot(
  aicc_data,
  aes(x = Ecotype, y = Baseline, color = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(
    aes(shape = Treatment),
    position = position_jitterdodge(jitter.width = 0.12, dodge.width = 0.75),
    size = 2.5,
    alpha = 0.8
  ) +
  scale_color_manual(values = pal_treatment) +
  scale_shape_manual(values = treatment_shape_map) +
  facet_wrap(~ Lake_type_label) +
  labs(
    title = "Change in fanning time by ecotype, treatment, and lake type",
    x = "Ecotype",
    y = "change in fanning time (s)",
    color = "Treatment",
    shape = "Treatment"
  ) +
  theme_paper()

print(baseline_ecotype_laketype_box)

# diagnostics/ check residuals 
baseline_resid <- resid(base_lmer3)
baseline_fitted <- fitted(base_lmer3)

hist(
  baseline_resid,
  breaks = 15,
  main = "Histogram of residuals: baseline model",
  xlab = "Residuals",
  col = "grey80",
  border = "white"
)

qqnorm(baseline_resid, main = "QQ plot of residuals: baseline model")
qqline(baseline_resid, col = "red", lwd = 2)



plot(
  baseline_fitted, baseline_resid,
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs fitted: baselinel model",
  pch = 16,
  col = "black"
)

abline(h = 0, lty = 2, col = "red")

print(shapiro.test(baseline_resid))

```

## AICc Model Comparison: Difference in time fanning


```{r}


# behavior model set
# all models use same dataset
# each model is biologically meaningful and each model has lake type and treatment as those two effects are crucial for the main question

Bigmodel <- lmer(
  Fan_dif ~ Treatment * Lake_type + Bodycondition + Nest_depth +
    Percent_cover + Latency_to_return + Embryos_stage_simple + (1 | Lake),
  data = aicc_data,
  REML = FALSE
)

model2 <- lmer(
  Fan_dif ~ Treatment * Lake_type + Nest_depth + Percent_cover +
    Latency_to_return + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model3 <- lmer(
  Fan_dif ~ Treatment * Lake_type + Bodycondition + Percent_cover +
    Latency_to_return + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model4 <- lmer(
  Fan_dif ~ Treatment * Lake_type + Bodycondition + Nest_depth +
    Latency_to_return + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model5 <- lmer(
  Fan_dif ~ Treatment * Lake_type + Bodycondition + Nest_depth +
    Percent_cover + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model6 <- lmer(
  Fan_dif ~ Treatment * Lake_type + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model7 <- lmer(
  Fan_dif ~ Treatment * Lake_type + Nest_depth + Percent_cover +
    Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model8 <- lmer(
  Fan_dif ~ Treatment + Lake_type + Bodycondition + Nest_depth +
    Percent_cover + Latency_to_return + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model9 <- lmer(
  Fan_dif ~ Treatment + Lake_type + Nest_depth + Percent_cover +
    Latency_to_return + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model10 <- lmer(
  Fan_dif ~ Treatment + Lake_type + Bodycondition + Nest_depth +
    Percent_cover + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model11 <- lmer(
  Fan_dif ~ Treatment + Lake_type + Nest_depth + Percent_cover +
    Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model12 <- lmer(
  Fan_dif ~ Treatment + Lake_type + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model13 <- lmer(
  Fan_dif ~ Treatment * Lake_type + (1 | Lake),
  data = aicc_data, REML = FALSE
)

model14 <- lmer(
  Fan_dif ~ Treatment + Lake_type + Embryos_stage_simple + (1 | Lake),
  data = aicc_data, REML = FALSE
)
  
  model15 <- lmer(
  Fan_dif ~ Treatment + Lake_type + Nest_depth + (1 | Lake),
  data = aicc_data, REML = FALSE
  )
  
  model16 <- lmer(
    Fan_dif ~1 + (1 | Lake),
  data = aicc_data,
  REML = FALSE
)

behavior_model_list <- list(
  Bigmodel = Bigmodel,
  model2 = model2,
  model3 = model3,
  model4 = model4,
  model5 = model5,
  model6 = model6,
  model7 = model7,
  model8 = model8,
  model9 = model9,
  model10 = model10,
  model11 = model11,
  model12 = model12,
  model13 = model13,
  model14 = model14,
  model15 = model15,
  model16 = model16
  
)

#Double check each model has same n
print(sapply(behavior_model_list, nobs))

aic_table <- MuMIn::model.sel(behavior_model_list) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Model_code") %>%
  dplyr::mutate(
    Formula = vapply(
      behavior_model_list,
      function(x) {
        f <- paste(deparse(stats::formula(x)), collapse = " ")
        f <- gsub("\\s+", " ", f)
        f
      },
      character(1)
    )[Model_code]
  ) %>%
  dplyr::select(Model_code, Formula, df, logLik, AICc, delta, weight) %>%
  dplyr::arrange(AICc)

print(
  knitr::kable(
    aic_table,
    digits = 3,
    escape = FALSE,
    caption = "AICc comparison of candidate behavioral models"
  ) %>%
    kableExtra::kable_styling(full_width = FALSE)
)

```

### Evaluating final selected model

```{r}


# final selected model from AICc
# refit winning model with REML = TRUE for final estimates

BestLMER_model <-  lmer(
  Fan_dif ~ Treatment + Lake_type + (1 | Lake),
  data = aicc_data, REML = TRUE
)

std_model <- standardize_parameters(
  BestLMER_model,
  method = "refit",
  ci = 0.95
)



# extract statistical outliers using IQR rule
outliers_fan <- boxplot.stats(aicc_data$Fan_dif)$out

# print outlier values
print(outliers_fan)

# show rows corresponding to outliers
aicc_data[aicc_data$Fan_dif %in% outliers_fan, ]

# number of outliers
length(outliers_fan)


# Make the exact model dataset
fan_model_data <- aicc_data %>%
  filter(
    !is.na(Fan_dif),
    !is.na(Treatment),
    !is.na(Lake_type),
    !is.na(Lake)
  ) %>%
  mutate(row_id = row_number())

# Fit full best model
bestLMER_model <- lmer(
  Fan_dif ~ Treatment + Lake_type + (1 | Lake),
  data = fan_model_data,
  REML = TRUE
)

summary(bestLMER_model)

# Identify the most extreme observations by absolute Fan_dif
extreme_fan <- fan_model_data %>%
  mutate(abs_Fan_dif = abs(Fan_dif)) %>%
  arrange(desc(abs_Fan_dif)) %>%
  select(row_id, Lake, Lake_type, Treatment, Fan_dif, abs_Fan_dif) %>%
  slice(1:5)

print(extreme_fan)

# Remove the single most extreme observation and refit
fan_no_top1 <- fan_model_data %>%
  filter(row_id != extreme_fan$row_id[1])

bestLMER_no_top1 <- lmer(
  Fan_dif ~ Treatment + Lake_type + (1 | Lake),
  data = fan_no_top1,
  REML = TRUE
)

summary(bestLMER_no_top1)

#  Remove the top 2 most extreme observations and refit
fan_no_top2 <- fan_model_data %>%
  filter(!row_id %in% extreme_fan$row_id[1:2])

bestLMER_no_top2 <- lmer(
  Fan_dif ~ Treatment + Lake_type + (1 | Lake),
  data = fan_no_top2,
  REML = TRUE
)

summary(bestLMER_no_top2)

# Compare fixed effects side by side
coef_full <- coef(summary(bestLMER_model)) %>%
  as.data.frame() %>%
  mutate(term = rownames(.), model = "Full data")

coef_no1 <- coef(summary(bestLMER_no_top1)) %>%
  as.data.frame() %>%
  mutate(term = rownames(.), model = "Top 1 removed")

coef_no2 <- coef(summary(bestLMER_no_top2)) %>%
  as.data.frame() %>%
  mutate(term = rownames(.), model = "Top 2 removed")

coef_compare <- bind_rows(coef_full, coef_no1, coef_no2) %>%
  select(model, term, Estimate, `Std. Error`, df, `t value`, `Pr(>|t|)`)

print(coef_compare)


bind_rows(
  tidy(bestLMER_model) %>% mutate(Model="Full"),
  tidy(bestLMER_no_top1) %>% mutate(Model="Top1 removed"),
  tidy(bestLMER_no_top2) %>% mutate(Model="Top2 removed")
) %>%
filter(term!="(Intercept)") %>%
select(Model, term, estimate, std.error, p.value)

# diagnostics/ check residuals 
best_resid <- resid(BestLMER_model)
best_fitted <- fitted(BestLMER_model)

hist(
  best_resid,
  breaks = 15,
  main = "Histogram of residuals: best behavioral model",
  xlab = "Residuals",
  col = "grey80",
  border = "white"
)

qqnorm(best_resid, main = "QQ plot of residuals: best behavioral model")
qqline(best_resid, col = "red", lwd = 2)



plot(
  best_fitted, best_resid,
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs fitted: best behavioral model",
  pch = 16,
  col = "black"
)

abline(h = 0, lty = 2, col = "red")

print(shapiro.test(best_resid))

##Let's look at the model now 
print(summary(BestLMER_model))
print(performance::check_collinearity(BestLMER_model))
print(MuMIn::r.squaredGLMM(BestLMER_model))
print(confint(BestLMER_model, method = "profile"))


# fixed effects table
behavior_fixed_effects <- broom.mixed::tidy(
  BestLMER_model,
  effects = "fixed",
  conf.int = TRUE
)

print(behavior_fixed_effects)

#standardrized effects for the best behavior model
behavior_std_effects <- effectsize::standardize_parameters(BestLMER_model)
print(behavior_std_effects)
MuMIn::r.squaredGLMM(BestLMER_model)

behavior_results_table <- behavior_fixed_effects %>%
  dplyr::mutate(
    term = dplyr::recode(
      term,
      `(Intercept)` = "Intercept",
      `TreatmentPike` = "Treatment: Pike vs Char",
      `TreatmentTrout` = "Treatment: Trout vs Char",
      `Lake_typePike` = "Lake type: Native + Invasive vs Native only",
  
    )
  ) %>%
  dplyr::select(term, estimate, std.error, statistic, df, p.value, conf.low, conf.high) %>%
  dplyr::rename(
    Term = term,
    Estimate = estimate,
    SE = std.error,
    t_value = statistic,
    DF = df,
    P_value = p.value,
    CI_low = conf.low,
    CI_high = conf.high
  )

print(
  knitr::kable(
    behavior_results_table,
    digits = 3,
    caption = "Fixed effects from the selected behavioral mixed-effects model"
  ) %>%
    kableExtra::kable_styling(full_width = FALSE)
)

# let's try estimated marginal meansn for our treatments 
emmeans_treat_main <- emmeans::emmeans(BestLMER_model, ~ Treatment)
emmeans_treat_pairs <- summary(pairs(emmeans_treat_main, adjust = "tukey"))
print(emmeans_treat_pairs)



aicc_data %>%
  group_by(Treatment) %>%
  summarise(
    n = n(),
    mean = mean(Latency_to_return),
    sd = sd(Latency_to_return),
    p = t.test(Latency_to_return, mu = 0)$p.value,
    CI.low = t.test(Latency_to_return, mu = 0)$conf.int[1],
    CI.high = t.test(Latency_to_return, mu = 0)$conf.int[2]
  )

```

### Visualization

```{r}


# behavior figures
# main boxplot
p_beh_lake_treat <- ggplot(
  aicc_data %>% dplyr::filter(!is.na(Lake_type_label), !is.na(Treatment), !is.na(Fan_dif)),
  aes(x = Lake_type_label, y = Fan_dif, fill = Treatment)
) +
  geom_boxplot(
    aes(group = interaction(Lake_type_label, Treatment)),
    outlier.shape = NA,
    width = 0.65,
    alpha = 0.85,
    position = position_dodge(width = 0.75)
  ) +
  geom_point(
    aes(shape = Treatment, group = interaction(Lake_type_label, Treatment)),
    color = "black",
    size = 2.4,
    alpha = 0.85,
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.75
    )
  ) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  scale_fill_manual(values = pal_treatment) +
  scale_shape_manual(values = treatment_shape_map) +
  labs(
    title = "Difference in fanning time by lake type and predator treatment",
    x = "Lake type",
    y = "Difference in fanning time(s)",
    fill = "Treatment",
    shape = "Treatment"
  ) +
  theme_paper()

print(p_beh_lake_treat)

# fixed-effect coefficient plot
coef_plot_df <- behavior_fixed_effects %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    term = dplyr::recode(
      term,
      `TreatmentPike` = "Treatment: Pike vs Char",
      `TreatmentTrout` = "Treatment: Trout vs Char",
      `Lake_typePike` = "Lake type: Native + Invasive vs Native only",
     
    )
  )

p_coef <- ggplot(
  coef_plot_df,
  aes(x = estimate, y = reorder(term, estimate))
) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.15,
    linewidth = 0.7
  ) +
  geom_point(size = 3) +
  labs(
    title = "Fixed effects from the best behavioral model",
    x = "Estimate (95% CI)",
    y = NULL
  ) +
  theme_paper()

print(p_coef)

# predicted marginal means plot
emm_plot_df <- as.data.frame(
  emmeans::emmeans(BestLMER_model, ~ Treatment + Lake_type)
) %>%
  dplyr::mutate(
    Lake_type_label = dplyr::case_when(
      Lake_type == "No Pike" ~ "Native predator only lakes",
      Lake_type == "Pike" ~ "Native + Invasive predator lakes",
      TRUE ~ as.character(Lake_type)
    )
  )

p_emm <- ggplot(
  emm_plot_df,
  aes(x = Treatment, y = emmean, color = Treatment, group = 1)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.8) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.12,
    linewidth = 0.8
  ) +
  facet_wrap(~ Lake_type_label) +
  scale_color_manual(values = pal_treatment) +
  labs(
    title = "Estimated marginal means of predator-induced change in fanning time",
    x = "Treatment",
    y = "Estimated mean change in fanning time (s)",
    color = "Treatment"
  ) +
  theme_paper()

print(p_emm)

# reaction norm: raw means
raw_means_df <- aicc_data %>%
  dplyr::group_by(Lake_type_label, Treatment) %>%
  dplyr::summarise(
    mean_Fan_dif = mean(Fan_dif, na.rm = TRUE),
    sd_Fan_dif = sd(Fan_dif, na.rm = TRUE),
    n = dplyr::n(),
    se_Fan_dif = sd_Fan_dif / sqrt(n),
    .groups = "drop"
  )

p_reaction_raw <- ggplot(
  raw_means_df,
  aes(x = Treatment, y = mean_Fan_dif, color = Lake_type_label, group = Lake_type_label)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.9) +
  geom_errorbar(
    aes(ymin = mean_Fan_dif - se_Fan_dif, ymax = mean_Fan_dif + se_Fan_dif),
    width = 0.12,
    linewidth = 0.7
  ) +
  scale_color_manual(values = pal_lake) +
  labs(
    title = "Reaction norm plot of raw mean change in fanning time",
    x = "Treatment",
    y = "Mean change in fanning time (s)",
    color = "Lake type"
  ) +
  theme_paper()

print(p_reaction_raw)

# reaction norm: model means
p_reaction_emm <- ggplot(
  emm_plot_df,
  aes(x = Treatment, y = emmean, color = Lake_type_label, group = Lake_type_label)
) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.9) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.12,
    linewidth = 0.7
  ) +
  scale_color_manual(values = pal_lake) +
  labs(
    title = "Reaction norm plot of model-estimated mean change in fanning time",
    x = "Treatment",
    y = "Estimated mean change in fanning time (s)",
    color = "Lake type"
  ) +
  theme_paper()

print(p_reaction_emm)

plot_data <- aicc_data %>%
  filter(!is.na(Lake_type_label),
         !is.na(Treatment),
         !is.na(Fan_dif))

summary_fan <- plot_data %>%
  group_by(Lake_type_label, Treatment) %>%
  summarise(
    n = n(),
    mean = mean(Fan_dif, na.rm = TRUE),
    se = sd(Fan_dif, na.rm = TRUE) / sqrt(n),
    ci = qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

p_fan_lake_treat <- ggplot(
  plot_data,
  aes(x = Lake_type_label,
      y = Fan_dif,
      fill = Treatment,
      color = Treatment)
) +
  geom_hline(yintercept = 0, linetype = "dashed",
             color = "grey55", linewidth = 0.5) +

  # boxplot
  geom_crossbar(
    data = plot_data %>%
      group_by(Lake_type_label, Treatment) %>%
      summarise(
        q1 = quantile(Fan_dif, 0.25, na.rm = TRUE),
        med = median(Fan_dif, na.rm = TRUE),
        q3 = quantile(Fan_dif, 0.75, na.rm = TRUE),
        .groups = "drop"
      ),
    aes(
      x = Lake_type_label,
      y = med,
      ymin = q1,
      ymax = q3,
      group = Treatment,
      fill = Treatment,
      color = Treatment
    ),
    position = position_dodge(width = 0.78),
    width = 0.32,
    alpha = 0.18,
    linewidth = 0.45
  ) +

  # individual males
  geom_point(
    aes(shape = Treatment),
    position = position_jitterdodge(jitter.width = 0.07,
                                    dodge.width = 0.78),
    size = 2.7,
    alpha = 0.9
  ) +

  # mean ± 95% CI
  geom_pointrange(
    data = summary_fan,
    aes(x = Lake_type_label,
        y = mean,
        ymin = mean - ci,
        ymax = mean + ci,
        group = Treatment),
    inherit.aes = FALSE,
    position = position_dodge(width = 0.78),
    shape = 18,
    size = 0.25,
    linewidth = 0.55,
    color = "black"
  ) +

  scale_fill_manual(values = pal_treatment) +
  scale_color_manual(values = pal_treatment) +
  scale_shape_manual(values = treatment_shape_map) +

  labs(
    title = "Change in fanning time by lake type and predator treatment",
    x = "Lake type",
    y = "Difference in fanning time (s)",
    fill = "Treatment",
    color = "Treatment",
    shape = "Treatment"
  ) +
  theme_paper()

print(p_fan_lake_treat)

ggsave(
  filename = "Difference_fanning_time.png",
  plot = p_fan_lake_treat, 
  width = 8,
  height = 6,
  units = "in",
  dpi = 600,
  bg = "white"
)

###return to nest data
	ggplot(aicc_data,
       aes(x = Treatment,
           y = Latency_to_return,
           color = Treatment,
           fill = Treatment,
           shape = Treatment)) +


  geom_boxplot(width = 0.18,
               alpha = 0.25,
               outlier.shape = NA,
               linewidth = 0.5) +

  geom_jitter(width = 0.08,
              size = 2.2,
              alpha = 0.85) +

  stat_summary(fun = mean,
               geom = "point",
               shape = 18,
               size = 2.5,
               color = "black") +

  geom_hline(yintercept = 0,
             linetype = "dashed") +

  labs(
    x = "Predator treatment",
    y = "Latency to return to nest (s)"
  ) +

  theme_paper()



ggplot(aicc_data, aes(x = Lake_type_label, y = Baseline, fill = Lake_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6, width = 0.6) +
  geom_jitter(width = 0.25, height = 0, size = 2.5, color = "black", alpha = 0.8) +
  labs(x = "Lake type", y = "Baseline fanning time (s)") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

	
```


#Transcriptomic data

## Overview

Analyze whole brain gene expression between predator communities, and between parental state (parent vs non-parent)

## Load necessary libraries and data

```{r}

shelf(genefilter, DESeq2, apeglm, vsn, tidyr, RColorBrewer, dplyr, tidyverse, ggrepel, ggplot2, stringr, AnnotationDbi, clusterProfiler, enrichplot, forcats, GO.db, biomaRt, tibble, pheatmap)

#######
# Loading in raw data and meta data

akdata <- read.csv("AK24_Predator_GeneCounts.csv", check.names = FALSE)
akmeta <- read.csv("AK24_Predator_metadata.csv", check.names = FALSE, stringsAsFactors = FALSE)

```

## Checking and cleaning data as necessary

```{r}

# Confirming that the data files are a matrix of count data.

akdata <- data.frame(akdata[, -1], row.names = akdata[, 1])

if ("sample" %in% names(akmeta)) {
  rownames(akmeta) <- akmeta$sample
} else {
  rownames(akmeta) <- akmeta[, 1]
  akmeta <- akmeta[, -1, drop = FALSE]
}

head(akdata)
head(akmeta)

# Make sure the sample names are in the same order between the two files.

all(colnames(akdata) %in% rownames(akmeta))
akdata <- akdata[, rownames(akmeta), drop = FALSE]
all(colnames(akdata) == rownames(akmeta))

# Cleaning metadata and setting factor levels

akmeta$Batch <- factor(akmeta$Batch)

akmeta$Lake_type <- trimws(as.character(akmeta$Lake_type))
akmeta$Lake_type[akmeta$Lake_type %in% c("No Pike", "No pike", "no pike")] <- "No Pike"
akmeta$Lake_type[akmeta$Lake_type %in% c("Pike", "pike")] <- "Pike"
akmeta$Lake_type <- factor(akmeta$Lake_type, levels = c("No Pike", "Pike"))

akmeta$Treatment <- trimws(as.character(akmeta$Treatment))
akmeta$Treatment <- factor(akmeta$Treatment, levels = c("Char", "Pike", "Trout"))

akmeta$Parent_state <- trimws(as.character(akmeta$Parent_state))
akmeta$Parent_state[akmeta$Parent_state %in% c("Parenting", "parenting")] <- "Parenting"
akmeta$Parent_state[akmeta$Parent_state %in% c("Non-parenting", "Non parenting", "non-parenting", "non parenting")] <- "Non-parenting"
akmeta$Parent_state <- factor(akmeta$Parent_state, levels = c("Non-parenting", "Parenting"))

akmeta$Embryo_stage <- trimws(as.character(akmeta$Embryo_stage))
akmeta$Embryo_stage[akmeta$Embryo_stage == ""] <- NA
akmeta$Embryo_stage <- factor(akmeta$Embryo_stage)

akmeta$Treatment <- relevel(akmeta$Treatment, ref = "Char")
akmeta$Lake_type <- relevel(akmeta$Lake_type, ref = "No Pike")
akmeta$Parent_state <- relevel(akmeta$Parent_state, ref = "Non-parenting")

akmeta$Lake_type_label <- ifelse(
  akmeta$Lake_type == "No Pike",
  "Native predator-only",
  "Native + invasive predator"
)
akmeta$Lake_type_label <- factor(
  akmeta$Lake_type_label,
  levels = c("Native predator-only", "Native + invasive predator")
)

# Plotting histogram and mean-variance relationship

row_mean1 <- apply(akdata, 1, mean)
row_var1 <- apply(akdata, 1, var)
df <- data.frame(row_mean1, row_var1)

hist(log(df$row_mean1 + 1))
plot(log(df$row_mean1 + 1), log(df$row_var1 + 1), abline(0,1), ylim = c(0,30), xlim = c(0,30))

```

## Filtering genes and running analyses

```{r}


# Creating dds object with batch, lake type, and parent state in the model

dds <- DESeqDataSetFromMatrix(countData = akdata, colData = akmeta, design = ~ Batch + Lake_type + Parent_state)

# Pre-filtering out genes with less than 10 counts in fewer than 3 samples

keep <- rowSums(counts(dds) >= 10) >= 3
dds <- dds[keep, ]
names(dds)

# Running the analysis with Wald test, followed by shrinkage of log2 fold change to aid visualization

dds <- DESeq(dds)
res <- results(dds, alpha = 0.05)
resultsNames(dds)

# Variance stabilizing transformation to improve clustering visualization

vsd <- vst(dds, blind = FALSE)

```

## Visualization

```{r}

# Plotting vsd plots, before and after vst transformation

ntd <- normTransform(dds)
meanSdPlot(assay(ntd))
meanSdPlot(assay(vsd))

# PC1 and PC2 for Lake type, Batch, Parent state, and Parent state + Lake type

plotPCA(vsd, intgroup = "Lake_type")
plotPCA(vsd, intgroup = "Batch")
plotPCA(vsd, intgroup = "Parent_state")
plotPCA(vsd, intgroup = c("Parent_state", "Lake_type"))

# Plotting PC1 and PC2 manually for Parent state and Lake type

pca_data <- plotPCA(vsd, intgroup = c("Parent_state", "Lake_type"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data) +
  geom_point(aes(x = PC1, y = PC2, color = Parent_state, shape = Lake_type), size = 3.2) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("Transcriptomic PCA") +
  theme_classic()

# Getting normalized counts

count_matrix <- counts(dds, normalized = TRUE)

# Making the results tables.

meta_table <- akmeta %>%
  rownames_to_column(var = "samplename") %>%
  as_tibble()

norm_counts <- count_matrix %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

```

## Results 


```{r}

# Setting thresholds for what constitutes a DEG.

padj.cutoff <- 0.05
lfc.cutoff <- 0.58

# Lake type results

lake_name <- resultsNames(dds)[grep("Lake_type", resultsNames(dds))]
res_lake <- results(dds, name = lake_name)
resLFC_lake <- lfcShrink(dds, coef = lake_name, type = "apeglm")
summary(resLFC_lake)

# Parent state results

parent_name <- resultsNames(dds)[grep("Parent_state", resultsNames(dds))]
res_parent <- results(dds, name = parent_name)
resLFC_parent <- lfcShrink(dds, coef = parent_name, type = "apeglm")
summary(resLFC_parent)

# Batch results


batch_names <- resultsNames(dds)[grep("^Batch_", resultsNames(dds))]
res_batch <- lapply(batch_names, function(x) results(dds, name = x))
names(res_batch) <- batch_names
resLFC_batch <- lapply(batch_names, function(x) lfcShrink(dds, coef = x, type = "apeglm"))
names(resLFC_batch) <- batch_names
lapply(resLFC_batch, summary)

# MA Plots before and after shrinkage

plotMA(res_lake, ylim = c(-3,3))
plotMA(resLFC_lake, ylim = c(-3,3))

plotMA(res_parent, ylim = c(-3,3))
plotMA(resLFC_parent, ylim = c(-3,3))

# Making res_table to eventually get DEGs

res_table_lake <- res_lake %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

res_table_parent <- res_parent %>%
  data.frame() %>%
  rownames_to_column(var = "gene") %>%
  as_tibble()

DEG_lake <- res_table_lake %>%
  filter(padj < padj.cutoff & abs(log2FoldChange) >= lfc.cutoff)

DEG_parent <- res_table_parent %>%
  filter(padj < padj.cutoff & abs(log2FoldChange) >= lfc.cutoff)

```

## Collecting and visualizing top 30 DEG by lake type

```{r}

# Getting the top 30 DEG from the lake-type res_table and subset those out of the normalized counts

top30_DEG_lake <- res_table_lake %>%
  arrange(padj) %>%
  pull(gene) %>%
  head(n = 30)

top30_DEG_norm_lake <- norm_counts %>%
  filter(gene %in% top30_DEG_lake)

# Converting from wide to long format to plot

top30_DEG_norm_long_lake <- top30_DEG_norm_lake %>%
  gather(colnames(top30_DEG_norm_lake)[2:ncol(top30_DEG_norm_lake)], key = "samplename", value = "normalized_counts")

top30_DEG_norm_long_lake <- inner_join(meta_table, top30_DEG_norm_long_lake, by = "samplename")

# Plotting top 30 DEGs

ggplot(top30_DEG_norm_long_lake) +
  geom_point(aes(x = gene, y = normalized_counts, color = Lake_type_label)) +
  scale_y_log10() +
  xlab("Genes") +
  ylab("log10 Normalized Counts") +
  ggtitle("Top 30 Significant DE Genes") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))

# Preparing for volcano plots - need to have a column in the results table stating whether or not the gene is considered differentially expressed

res_table_lake <- res_table_lake %>%
  mutate(threshold = padj < 0.05 & abs(log2FoldChange) >= 0.58)

res_table_parent <- res_table_parent %>%
  mutate(threshold = padj < 0.05 & abs(log2FoldChange) >= 0.58)

```


## Volcano Plots

```{r}

# Making the volcano plot showing the DEGs

ggplot(res_table_lake) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  scale_color_manual(values = c("blue", "green")) +
  geom_hline(yintercept = 1.30, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = c(-0.58,0.58), linetype = "dashed", color = "black") +
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = rel(1.5), hjust = 0.3),
    axis.title = element_text(size = rel(1.25))
  )

ggplot(res_table_parent) +
  geom_point(aes(x = log2FoldChange, y = -log10(padj), colour = threshold)) +
  scale_color_manual(values = c("blue", "green")) +
  geom_hline(yintercept = 1.30, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = c(-0.58,0.58), linetype = "dashed", color = "black") +
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = rel(1.5), hjust = 0.3),
    axis.title = element_text(size = rel(1.25))
  )

# Volcano plot showing the top 10 DEGs

res_table_lake <- res_table_lake %>% arrange(padj) %>% mutate(genelabels = "")
res_table_lake$genelabels[1:min(10, nrow(res_table_lake))] <- res_table_lake$gene[1:min(10, nrow(res_table_lake))]

ggplot(res_table_lake, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(colour = threshold)) +
  geom_text_repel(aes(label = genelabels)) +
  scale_color_manual(values = c("black", "red3")) +
  scale_y_continuous(expand = c(0, 0)) +
  xlab("log2 fold change") +
  ylab("-log10 adjusted p-value") +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = rel(1.5), hjust = 0.5),
    axis.title = element_text(size = rel(1.25))
  )

```

## Summaries of DEGs

```{r}

#  summaries 

count_deg <- function(res_obj, padj_cutoff = 0.05, lfc_cutoff = 0.58) {
  data.frame(
    padj_only = sum(!is.na(res_obj$padj) & res_obj$padj < padj_cutoff),
    padj_and_lfc = sum(!is.na(res_obj$padj) &
                         res_obj$padj < padj_cutoff &
                         abs(res_obj$log2FoldChange) >= lfc_cutoff)
  )
}

deg_summary <- bind_rows(
  count_deg(res_batch[[1]]) %>% mutate(Effect = names(res_batch)[1]),
  count_deg(res_batch[[2]]) %>% mutate(Effect = names(res_batch)[2]),
  count_deg(res_batch[[3]]) %>% mutate(Effect = names(res_batch)[3]),
  count_deg(res_lake) %>% mutate(Effect = "Lake type"),
  count_deg(res_parent) %>% mutate(Effect = "Parent state")
  ) %>%
  dplyr::select(Effect, padj_only, padj_and_lfc)

print(deg_summary)

```


## Heatmap of lake-type DEG

```{r}


# Heatmap of top 30 lake-type DEGs

deg_ids <- as.data.frame(res_lake) %>%
  rownames_to_column("gene") %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) >= 0.58) %>%
  arrange(padj) %>%
  slice_head(n = 30) %>%
  pull(gene)

deg_ids <- deg_ids[deg_ids %in% rownames(vsd)]

mat <- assay(vsd)[deg_ids, , drop = FALSE]

meta_heat <- akmeta[match(colnames(mat), rownames(akmeta)), , drop = FALSE]

meta_heat$Lake_type_label <- ifelse(
  meta_heat$Lake_type %in% c("No Pike", "No pike"),
  "Native predator-only",
  "Native + invasive predator"
)

meta_heat$Lake_type_label <- factor(
  meta_heat$Lake_type_label,
  levels = c("Native predator-only", "Native + invasive predator")
)

annotation_col <- data.frame(
  Lake_type = meta_heat$Lake_type_label
)
rownames(annotation_col) <- colnames(mat)

ann_colors <- list(
  Lake_type = c(
    "Native predator-only" = "#00BFC4",
    "Native + invasive predator" = "#F8766D"
  )
)

pheatmap(
  mat,
  scale = "row",
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  show_rownames = FALSE,
  fontsize_col = 10,
  main = "Top lake-type DEGs"
)

```

## PCAs by Lake Type

```{r}

# PCA by lake

plotPCA(vsd, intgroup = "Lake")

pca_lake <- plotPCA(vsd, intgroup = c("Lake"), returnData = TRUE)
percentVar_lake <- round(100 * attr(pca_lake, "percentVar"))

ggplot(pca_lake, aes(PC1, PC2, color = Lake, label = name)) +
  geom_point(size = 3) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA by lake",
    x = paste0("PC1: ", percentVar_lake[1], "% variance"),
    y = paste0("PC2: ", percentVar_lake[2], "% variance")
  )

# GO/GSEA for lake-type contrast
##requires server connection, im going to make a csv with gene annotations and use this to help with reproducibility
#mart <- biomaRt::useEnsembl(
 # biomart = "genes",
  #dataset = "gaculeatus_gene_ensembl"
#)

all_genes <- rownames(dds)

#go_map <- biomaRt::getBM(
 # attributes = c("ensembl_gene_id", "go_id"),
  #filters = "ensembl_gene_id",
  #values = all_genes,
  #mart = mart
#) %>%
 # filter(go_id != "") %>%
  #distinct()

#TERM2GENE <- go_map %>%
  #dplyr::select(go_id, ensembl_gene_id)

#TERM2NAME <- AnnotationDbi::select(
 # GO.db,
  #keys = unique(TERM2GENE$go_id),
  #keytype = "GOID",
  #columns = c("TERM")
#) %>%
 # distinct(GOID, TERM) %>%
  #rename(go_id = GOID, term_name = TERM)

annotations <- read.csv("annotations.csv")
TERM2GENE <- read.csv("TERM2GENE.csv")
TERM2NAME <- read.csv("TERM2NAME.csv")

# Ranked gene list for GSEA

gene_list_lake <- res_lake$stat
names(gene_list_lake) <- rownames(res_lake)
gene_list_lake <- gene_list_lake[!is.na(gene_list_lake)]
gene_list_lake <- sort(gene_list_lake, decreasing = TRUE)

gsea_lake <- clusterProfiler::GSEA(
  geneList = gene_list_lake,
  TERM2GENE = TERM2GENE,
  TERM2NAME = TERM2NAME,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  verbose = FALSE
)

gsea_lake_df <- as.data.frame(gsea_lake)
head(gsea_lake_df)

```

## Dotplot and Ridgeplots

```{r}

# Dotplot and ridgeplot

dotplot(gsea_lake, showCategory = 15)

ridgeplot(gsea_lake, showCategory = 10) +
  ggtitle("GSEA ridgeplot: lake-type differences") +
  labs(
    x = "Higher in native predator-only  <---- ranked gene statistic ---->  Higher in native + invasive predator lakes",
    y = NULL
  ) +
  theme_classic(base_size = 13)

# GO enrichment using only strong lake-type DEGs

lake_deg_ids <- DEG_lake$gene

ego_lake <- clusterProfiler::enricher(
  gene = lake_deg_ids,
  universe = all_genes,
  TERM2GENE = TERM2GENE,
  TERM2NAME = TERM2NAME,
  pAdjustMethod = "BH"
)

if (!is.null(ego_lake) && nrow(as.data.frame(ego_lake)) > 0) {
  dotplot(ego_lake, showCategory = 15)
  head(as.data.frame(ego_lake))
} else {
  message("No significant DEG-based GO enrichment found for lake-type DEGs.")
}

#  GSEA for parent state

gene_list_parent <- res_parent$stat
names(gene_list_parent) <- rownames(res_parent)
gene_list_parent <- gene_list_parent[!is.na(gene_list_parent)]
gene_list_parent <- sort(gene_list_parent, decreasing = TRUE)

gsea_parent <- clusterProfiler::GSEA(
  geneList = gene_list_parent,
  TERM2GENE = TERM2GENE,
  TERM2NAME = TERM2NAME,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  verbose = FALSE
)

gsea_parent_df <- as.data.frame(gsea_parent)
head(gsea_parent_df)

if (nrow(gsea_parent_df) > 0) {
  dotplot(gsea_parent, showCategory = 15)
}

# Quick summary lines

cat("\nDEG summary:\n")
print(deg_summary)

cat("\nNumber of lake-type DEGs:\n")
print(nrow(DEG_lake))

cat("\nNumber of parent-state DEGs:\n")
print(nrow(DEG_parent))

```


# Plotting the sites

```{r}

#Plotting the sites!

shelf(ggplot2, dplyr, sf, ggrepel, ggspatial, prettymapr, rnaturalearth, rnaturalearthdata)


# lake coordinates
lakes_map <- data.frame(
  Lake = c("South Rolly", "Crystal", "Finger", "Watson", "Hope", "Spirit"),
  Lake_type = c(
    "Native + invasive predator",
    "Native + invasive predator",
    "Native + invasive predator",
    "Native predators only",
    "Native predators only",
    "Native predators only"
  ),
  Longitude = c(
    -150.135689, -150.102865, -149.279200,
    -150.467000, -151.188278, -150.997664
  ),
  Latitude = c(
    61.667545, 61.710514, 61.605600,
    60.539000, 60.421528, 60.596104
  )
)

pal_lake_map <- c(
  "Native predators only" = "#0072B2",
  "Native + invasive predator" = "#CC79A7"
)

# convert lakes to sf object
lakes_sf <- st_as_sf(
  lakes_map,
  coords = c("Longitude", "Latitude"),
  crs = 4326
)

# transform to Web Mercator (required for map tiles)
lakes_sf_3857 <- st_transform(lakes_sf, 3857)

# plot
p_lake_map <- ggplot() +

  annotation_map_tile(
    type = "osm",
    zoom = 7
  ) +

  geom_sf(
    data = lakes_sf_3857,
    aes(color = Lake_type),
    size = 4
  ) +

  geom_sf_text(
    data = lakes_sf_3857,
    aes(label = Lake),
    size = 3.8,
    fontface = "bold",
    nudge_y = 15000
  ) +

  scale_color_manual(values = pal_lake_map) +

  coord_sf(
    crs = 3857,
    xlim = c(-17000000, -16500000),
    ylim = c(8400000, 8900000),
    expand = FALSE
  ) +

  # Scale bar
  annotation_scale(
    location = "bl",
    width_hint = 0.25,
    unit_category = "metric",
    text_cex = 0.8,
    line_width = 0.8
  ) +

  # North arrow
  annotation_north_arrow(
    location = "tr",
    which_north = "true",
    style = north_arrow_fancy_orienteering
  ) +

  labs(
    title = "Study lakes in southcentral Alaska",
    x = "Longitude",
    y = "Latitude",
    color = "Lake type"
  ) +

  theme_bw(base_size = 14) +

  theme(
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 16
    ),
    axis.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_line(
      color = "grey70",
      linewidth = 0.3
    )
  )

print(p_lake_map)

ggsave(
  filename = "study_lakes_map.png",
  plot = p_lake_map,
  width = 8,
  height = 6,
  units = "in",
  dpi = 600,
  bg = "white"
)


```
```

# Session Information

```{r}

sessionInfo()

```